Marginally-Stable Thermal Equilibria of Rayleigh-Bénard Convection

### 2021-04-09 publication_materials/ms_rbc.tex
sections 2 and 3: rev 2a

Aggressive convection is associated with large Rayleigh numbers $Ra$ and is difficult to simulate. State of the art convection simulations performed by \cite{Zhu} have reached $Ra = 10^{14}$ but estimates for stars are $Ra \sim 10^{20}$ \cite{Ossendrijver}. 
The scaling behavior of the Nusselt number $\Nu$ (convective strength) in the highly turbulent ultimate regime is of particular interest.
There exists a substantial body of work pertaining to this specific topic with no general consensus \cite{Malkus}, \cite{Kraichnan} \cite{Spiegel}, \cite{Castaing}, \cite{Grossman}, \cite{Ahlers}.

@@

\section{Conventional Boussinesq Linear Stability Analysis}\label{sec:boussinesq}
We begin with the non-dimensionalized Boussinesq approximation for Rayleigh-Bénard Convection. 
The domain is 2-dimensional rectangular and horizontally periodic with spatial dimensions $0 < x < 4$ and $-1/2 < z < 1/2$. 
The fluid of interest is constrained between two planar boundaries at $z = -1/2$ and $z = 1/2$ with fixed temperatures $1$ and $0$ respectively. 
At both boundaries we specify impenetrable, no-slip conditions, such that the velocity $\mathbf{u} = u \hat{x} + w \hat{z} = \mathbf{0}$ at $z = \pm 1/2$. 
The equations of motion are then given by
\begin{align}
    \nabla \cdot \mathbf{u} &= 0 \label{EQ:motion1}\\
    \frac{\partial \mathbf{u}}{\partial t} + \mathbf{u} \cdot \nabla \mathbf{u} &= - \nabla p + T \hat{z} + \sqrt{\frac{Pr}{Ra}} \nabla^2 \mathbf{u} \label{EQ:motion2}\\
    \frac{\partial T}{\partial t} + \mathbf{u} \cdot \nabla T &= \frac{1}{\sqrt{RaPr}} \nabla^2 T \label{EQ:motion3}
\end{align}
where $p$ is pressure and $T$ is temperature. 
For completeness, we specify a final boundary condition $p = p_0$ at $z = \pm 1/2$. 
Any system of this form can be characterized by its Rayleigh number $Ra$ and Prandtl number $Pr$, where $Ra$ is the timescale ratio of thermal diffusion to convection and $Pr$ is the ratio of momentum diffusivity to thermal diffusivity.  
For convenience, we define 
\begin{equation}
\mathcal{R} = \sqrt{\frac{Pr}{Ra}}, \qquad \mathcal{P} = \frac{1}{\sqrt{Pr Ra}}.
\end{equation}

To derive the linearized system, we posit that each field can be represented as the sum of a mean profile (denoted by $\bar{\cdot } \;$) and a perturbation function (denoted by $\cdot'$). 
At an arbitrary time $t_0$, suppose $\frac{\partial \bar{T}}{\partial z} (z)$ is some given function which satisfies
\begin{equation}
    \int_{-\frac{1}{2}}^{\frac{1}{2}} \frac{\partial \bar{T}}{\partial z}\Big|_{t=t_0} dz = -1.
\end{equation}
Then the mean profiles are given by
\begin{align}
    \mathbf{\bar{u}} &= \mathbf{0} \\
    \bar{T}(z) &= 1 + \int_{-\frac{1}{2}}^{z} \frac{\partial \bar{T}}{\partial \zeta}\Big|_{t=t_0}(\zeta) d\zeta \\
    \bar{p}(z) &= p_0 + \int_{-\frac{1}{2}}^z \bar{T}\Big|_{t=t_0}(\zeta) d\zeta.
\end{align}
Accordingly, the fields are defined as
\begin{align}
    \mathbf{u}(x, z, t) &= \mathbf{u'}(x, z, t) \label{EQ:reynolds_dc_u}\\
    &= u'(x, z, t)\hat{x} + w'(x, z, t)\hat{z} \\
    T(x, z, t) &= \bar{T}(z) + T'(x, z, t) \label{EQ:reynolds_dc_T}\\
    p(x, z, t) &= \bar{p}(z) + p'(x, z, t).
\end{align}
Upon substitution, we equate terms which scale linearly with the perturbations, yielding the linearized stability equations
\begin{align}
    \nabla \cdot \mathbf{u'} &= 0 \label{EQ:linear1}\\
    \frac{\partial\mathbf{u'}}{\partial t} &= - \nabla p' + T'\hat{z} + \mathcal{R} \nabla^2 \mathbf{u'} \label{EQ:linear2}\\
    \frac{\partial T'}{\partial t} + \frac{\partial \bar{T}}{\partial z} w' &= \mathcal{P} \nabla^2 T' \label{EQ:linear3}
\end{align}
with Dirichlet boundary conditions 
\begin{equation}
    T'|_{z = \pm \frac{1}{2}} = 0, \quad u'|_{z = \pm \frac{1}{2}} = 0, \quad p'|_{z = \pm \frac{1}{2}} = 0.
\end{equation}
In 1916, Lord Rayleigh observed that (\ref{EQ:linear1}) and (\ref{EQ:linear3}) can be manipulated into a separable form with generalized solutions
\begin{align}
    w'(x, z, t) &= W(z) \, e^{i(k_xx-st)} \label{EQ:normal_modes1}\\ 
    u'(x, z, t) &= U(z) \, e^{i(k_xx-st)} \label{EQ:normal_modes2}\\ 
    T'(x, z, t) &= \theta(z) \, e^{i(k_xx-st)} \label{EQ:normal_modes3}\\ 
    p'(x, z, t) &= P(z) \, e^{i(k_xx-st)}\label{EQ:normal_modes4}
\end{align}
where $s = \sigma + i\omega$ and $k_x$ is constrained, by periodicity, to the countably infinite set (spectrum) of wavenumbers
\begin{align}
    k_x \in \big\{\frac{n\pi}{2} \, \big| \, n \in \mathbb{N}\big\}.
\end{align}
For each $k_x$, we can assess the stability of the perturbations by solving for the eigenvalue $s$, whose imaginary component $\omega$ plays the role of an exponential growth rate. 
Solution yields a finite set of eigenvalues, among which, that with maximum $\omega$ is assumed to be dominant. 
Positive eigenvalues indicate that the system is unstable to small disturbances of wavenumber $k_x$, while negative eigenvalues indicate stability. 
A complete linear stability analysis requires solution over the full spectrum of wavenumbers. 
The prototypical case is used to demonstrate that the critical Rayleigh number $Ra_c = 1708$ when $\frac{\partial \bar{T}}{\partial z} = -1$.

\section{Temperature profile evolution}\label{sec:evolution}
For this investigation, we allow $ \bar{T}$ to vary in $z$ and $t$. Substituting (\ref{EQ:reynolds_dc_T}) into (\ref{EQ:motion3}) yields
\begin{align}
    \frac{\partial \bar{T}}{\partial t} + \mathbf{u}' \cdot \nabla T' + \frac{\partial T'}{\partial t} + \frac{\partial \bar{T}}{\partial z} w' = \mathcal{P} \nabla^2 T' + \mathcal{P} \nabla^2 \bar{T}.
    \label{EQ:T't_xz}
\end{align}
Subtracting (\ref{EQ:linear3}) from (\ref{EQ:T't_xz}) eliminates several terms. 
We can replace the advective term $\mathbf{u}' \cdot \nabla T'$ with $\grad \cdot (\mathbf{u}'T')$ due to (\ref{EQ:linear1}). 
Taking the horizontal average of (\ref{EQ:T't_xz}), this advective term can be reduced to
\begin{align}
    \frac{1}{4}\int_0^4 \grad \cdot (\mathbf{u}'T') dx &= \frac{\partial }{\partial z}  \langle w'T' \rangle
\end{align}
where $\langle \cdot \rangle$ denotes the horizontal average. 
The remaining terms in (\ref{EQ:T't_xz}) are unaffected as they are independent of $x$. 
The eigenfunctions $W(z),\, U(z),\, \theta(z),$ and $P(z)$ do not admit any natural normalization, as (\ref{EQ:linear1}) - (\ref{EQ:linear3}) and their subsequent forms are linear with respect to the perturbations. 
For completeness, we provide the normalization 
\begin{equation}
    \int_{-\frac{1}{2}}^{\frac{1}{2}} \big| \theta (z) \big| dz = 1
\end{equation}
and assign each perturbation a yet-unknown amplitude $A \in \mathbb{R}$, thereby allowing us to reduce (\ref{EQ:T't_xz}) to the initial value problem (IVP)
\begin{equation}
    \frac{\partial \bar{T}}{\partial t} + A^2 \langle w'T' \rangle = \mathcal{P}  \frac{\partial^2 \bar{T}}{\partial z^2}, \label{EQ:T0_IVP}
\end{equation}
upon which the remainder of this investigation is centered. 
It is important to keep in mind that the forcing term $\langle w'T' \rangle$ is obtained by solving an eigenvalue problem (EVP) which itself involves $\frac{\partial \bar{T}}{\partial z}$. 
We also assume this term is composed of eigenfunctions belonging to the dominant mode. 
\par For various $Ra$ and fixed $Pr = 1$, we seek symmetric, marginally-stable temperature profiles $\bar{T}_{eq}(z)$ which are thermally equilibrated, thereby satisfying $\frac{\partial \bar{T}_{eq}}{\partial t} = 0$ according to (\ref{EQ:T0_IVP}). 
We employ the \texttt{Dedalus} pseudo-spectral python framework to solve the EVP outlined in Section \ref{sec:boussinesq} as well as the IVP (\ref{EQ:T0_IVP}) by representing each field with a dealiased Chebyshev polynomial basis. 
The necessary number of basis functions varies, as eigenfunctions and temperature profiles associated with large Rayleigh numbers admit increasingly small-scale features. 
We supplement \texttt{Dedalus} with the \texttt{Eigentools} package to obtain the eigenfunctions in (\ref{EQ:T0_IVP}).

@@

Nusselt numbers $\Nu$ of various solutions are given in Figure \ref{fig:nu_vs_ra}. 
The marginally-stable thermally-equilibrated states and numerical simulation data exhibit power-law relationships with $Ra$. 
More specifically, $Ra \sim \Nu^3$ which agrees with Figures \ref{fig:bl_ra}, \ref{fig:kx_marginals}, and \ref{fig:del_inv} and the scaling argument given by \cite{Malkus}. 
The nonlinear solutions, which maximize heat flux, are obtained \cite{Waleffe}. 
$\Nu$ of both equilibria exceed that of simulations, which might be due to the transitional behaviors outlined by \cite{Yalniz} inhibiting heat flux.

@@

\par The marginally-stable equilibria are associated with a combination of eigenfunctions whose average kinetic energy is significantly larger than the statistical steady state, as given by Figure \ref{fig:ke_sim}. 

>>>>

The simulation of aggressive convection, which is associated with large Rayleigh numbers $Ra$, is expensive. 
State of the art convection simulations performed by \cite{Zhu} have reached $Ra = 10^{14}$ but estimates for stars are $Ra \sim 10^{20}$ \cite{Ossendrijver}. 
The scaling behavior of the Nusselt number $\Nu$ (convective strength) in the highly turbulent ultimate regime is of particular interest.
There exists a substantial body of work pertaining to this specific topic with no general consensus \cite{Malkus}, \cite{Kraichnan} \cite{Spiegel}, \cite{Castaing}, \cite{Grossman}, \cite{Ahlers}.

@@

\section{Model Setup}\label{sec:model}
We begin with the non-dimensionalized Boussinesq approximation for Rayleigh-Bénard Convection. 
The domain is 2-dimensional rectangular and horizontally periodic with spatial dimensions $0 < x < 4$ and $-1/2 < z < 1/2$. 
The fluid of interest is constrained between two planar boundaries at $z = -1/2$ and $z = 1/2$ with fixed temperatures $1$ and $0$ respectively. 
At both boundaries we specify impenetrable, no-slip conditions, such that the velocity $\mathbf{u} = u \hat{x} + w \hat{z} = \mathbf{0}$ at $z = \pm 1/2$. 
The equations of motion are then given by
\begin{align}
    \nabla \cdot \mathbf{u} &= 0 \label{EQ:motion1}\\
    \frac{\partial \mathbf{u}}{\partial t} + \mathbf{u} \cdot \nabla \mathbf{u} &= - \nabla p + T \hat{z} + \mathcal{R} \nabla^2 \mathbf{u} \label{EQ:motion2}\\
    \frac{\partial T}{\partial t} + \mathbf{u} \cdot \nabla T &= \mathcal{P} \nabla^2 T \label{EQ:motion3}
\end{align}
where $p$ is pressure and $T$ is temperature. 
For completeness, we specify a final boundary condition $p = p_0$ at $z = \pm 1/2$. 
Any system of this form can be characterized by its dimensionless Rayleigh number $Ra = \frac{g\alpha L^3 \Delta T}{\nu \kappa}$ and Prandtl number $Pr = \frac{\nu}{\kappa}$, where $g, \, \alpha, \, L, \, \Delta T, \nu, \kappa$ are the gravitational acceleration, coefficient of thermal expansion, domain height, boundary temperature difference, kinematic viscosity, and thermal diffusivity respectively. 
\begin{equation}
\mathcal{R} = \sqrt{\frac{Pr}{Ra}}, \qquad \mathcal{P} = \frac{1}{\sqrt{Pr Ra}}.
\end{equation}

To derive the linearized system, we posit that each field can be represented as the sum of a mean profile (denoted by $\bar{\cdot } \;$) and a perturbation function (denoted by $\cdot'$).
\begin{align}
    \mathbf{u}(x, z, t) &= \mathbf{u'}(x, z, t) \label{EQ:reynolds_dc_u}\\
    &= u'(x, z, t)\hat{x} + w'(x, z, t)\hat{z} \\
    T(x, z, t) &= \bar{T}(z) + T'(x, z, t) \label{EQ:reynolds_dc_T}\\
    p(x, z, t) &= \bar{p}(z) + p'(x, z, t).
\end{align}
Upon substitution, we equate terms which scale linearly with the perturbations, yielding the linearized stability equations
\begin{align}
    \nabla \cdot \mathbf{u'} &= 0 \label{EQ:linear1}\\
    \frac{\partial\mathbf{u'}}{\partial t} &= - \nabla p' + T'\hat{z} + \mathcal{R} \nabla^2 \mathbf{u'} \label{EQ:linear2}\\
    \frac{\partial T'}{\partial t} + \frac{\partial \bar{T}}{\partial z} w' &= \mathcal{P} \nabla^2 T' \label{EQ:linear3}
\end{align}
with Dirichlet boundary conditions 
\begin{equation}
    T'|_{z = \pm \frac{1}{2}} = 0, \quad u'|_{z = \pm \frac{1}{2}} = 0, \quad p'|_{z = \pm \frac{1}{2}} = 0.
\end{equation}
In 1916, Lord Rayleigh observed that (\ref{EQ:linear1}) and (\ref{EQ:linear3}) can be manipulated into a separable form with generalized solutions
\begin{align}
    w'(x, z, t) &= W(z) \, e^{i(k_xx-st)} \label{EQ:normal_modes1}\\ 
    u'(x, z, t) &= U(z) \, e^{i(k_xx-st)} \label{EQ:normal_modes2}\\ 
    T'(x, z, t) &= \theta(z) \, e^{i(k_xx-st)} \label{EQ:normal_modes3}\\ 
    p'(x, z, t) &= P(z) \, e^{i(k_xx-st)}\label{EQ:normal_modes4}
\end{align}
where $s = \sigma + i\omega$ and $k_x$ is constrained, by periodicity, to the countably infinite set (spectrum) of wavenumbers
\begin{align}
    k_x \in \big\{\frac{n\pi}{2} \, \big| \, n \in \mathbb{N}\big\}.
\end{align}
For each $k_x$, we can assess the stability of the perturbations by solving for the eigenvalue $s$, whose imaginary component $\omega$ plays the role of an exponential growth rate. 
Solution yields a finite set of eigenvalues, among which, that with maximum $\omega$ is assumed to be dominant. 
Positive eigenvalues indicate that the system is unstable to small disturbances of wavenumber $k_x$, while negative eigenvalues indicate stability. 
A complete linear stability analysis requires solution over the full spectrum of wavenumbers. 
The prototypical case is used to demonstrate that the critical Rayleigh number $Ra_c = 1708$ when $\frac{\partial \bar{T}}{\partial z} = -1$.

For this investigation, we allow $\bar{T}$ to vary in $z$ and $t$. Substituting (\ref{EQ:reynolds_dc_T}) into (\ref{EQ:motion3}) yields
\begin{align}
    \frac{\partial \bar{T}}{\partial t} + \mathbf{u}' \cdot \nabla T' + \frac{\partial T'}{\partial t} + \frac{\partial \bar{T}}{\partial z} w' = \mathcal{P} \nabla^2 T' + \mathcal{P} \nabla^2 \bar{T}.
    \label{EQ:T't_xz}
\end{align}
Subtracting (\ref{EQ:linear3}) from (\ref{EQ:T't_xz}) eliminates several terms. 
We can replace the advective term $\mathbf{u}' \cdot \nabla T'$ with $\grad \cdot (\mathbf{u}'T')$ due to (\ref{EQ:linear1}). 
Taking the horizontal average of (\ref{EQ:T't_xz}), this advective term can be reduced to
\begin{align}
    \frac{1}{4}\int_0^4 \grad \cdot (\mathbf{u}'T') dx &= \frac{\partial }{\partial z}  \langle w'T' \rangle
\end{align}
where $\langle \cdot \rangle$ denotes the horizontal average. 
The remaining terms in (\ref{EQ:T't_xz}) are unaffected as they are independent of $x$. 
The eigenfunctions $W(z),\, U(z),\, \theta(z),$ and $P(z)$ do not admit any natural normalization, as (\ref{EQ:linear1}) -- (\ref{EQ:linear3}) and their subsequent forms are linear with respect to the perturbations. 
For completeness, we provide the normalization 
\begin{equation}
    \int_{-\frac{1}{2}}^{\frac{1}{2}} \big| \theta (z) \big| dz = 1
\end{equation}
and assign each perturbation a yet-unknown amplitude $A \in \mathbb{R}$, thereby allowing us to reduce (\ref{EQ:T't_xz}) to the initial value problem (IVP)
\begin{equation}
    \frac{\partial \bar{T}}{\partial t} + A^2 \langle w'T' \rangle = \mathcal{P}  \frac{\partial^2 \bar{T}}{\partial z^2}, \label{EQ:T0_IVP}
\end{equation}
upon which the remainder of this investigation is centered. 
It is important to keep in mind that the forcing term $\langle w'T' \rangle$ is obtained by solving an eigenvalue problem (EVP) which itself involves $\frac{\partial \bar{T}}{\partial z}$. 
We also assume this term is composed of eigenfunctions belonging to the dominant mode.

\section{Perturbation Evolution}
Solving the linearized system (\ref{EQ:linear1}) -- (\ref{EQ:linear3}) does not yield $A^2$. 
To evolve $\bar{T}$ numerically, we assume the perturbations and mean quantities depend on different time-scales, as in \cite{Chini_sw}. On the fast time-scale, stable modes will decay and unstable modes will act according to (\ref{EQ:T0_IVP}). 
We make the ansatz that only marginally stable modes will affect $\bar{T}$ on long time-scales
\begin{equation}
    \max_{k_x} \{ \omega \} = 0
\end{equation}
thereby constraining $A^2$ at each timestep.

For various $Ra$ and fixed $Pr = 1$, we seek symmetric, marginally-stable temperature profiles $\bar{T}_{eq}(z)$ which are thermally equilibrated, thereby satisfying $\frac{\partial \bar{T}_{eq}}{\partial t} = 0$ according to (\ref{EQ:T0_IVP}). 
We employ the \texttt{Dedalus} pseudo-spectral python framework to solve the EVP outlined in Section \ref{sec:model} as well as the IVP (\ref{EQ:T0_IVP}) by representing each field with a dealiased Chebyshev polynomial basis. 
The necessary number of basis functions varies, as eigenfunctions and temperature profiles associated with large Rayleigh numbers admit increasingly small-scale features. 
We supplement \texttt{Dedalus} with the \texttt{Eigentools} package to obtain the eigenfunctions in (\ref{EQ:T0_IVP}).

@@

Nusselt numbers $\Nu$ of various solutions are given in Figure \ref{fig:nu_vs_ra}. 
The marginally-stable thermally-equilibrated states and numerical simulation data exhibit power-law relationships with $Ra$. 
More specifically, $Ra \sim \Nu^3$ which agrees with Figures \ref{fig:bl_ra}, \ref{fig:kx_marginals}, and \ref{fig:del_inv} and the scaling argument given by \cite{Malkus}. 
The nonlinear solutions, which maximize heat flux, are obtained by \cite{Waleffe}. 
$\Nu$ of both equilibria exceed that of simulations, which might be due to the transitional behaviors outlined by \cite{Yalniz} and \cite{Cvitanovic} inhibiting heat flux. 
Alternatively, we might anticipate the existence of exact coherent states with smaller $\Nu$, forming another node in the Markov chain whose long-term behavior agrees with the simulation data.

@@

The marginally-stable equilibria are associated with a combination of eigenfunctions whose average kinetic energy is significantly larger than the statistical steady state, as given by Figure \ref{fig:ke_sim}. 
